Data Preparation, Integration and Cleaning

### 2017-02-04 abbDataPrep_Script.Rmd
finalizing data cleaning

---
title: "Data Preparation, Integration and Cleaning"
output:
  html_document: default
  html_notebook: default
---

@@

    data.path <- 'c:/dropbox/research/airBNB/data/'
    code.path <- 'c:/code/'

@@

    data.path <- 'alternate data path'
    code.path <- 'alternate data path'

@@

@@

  str.data <- read.csv(paste0(data.path, '/prepared/propdata.csv'), header=T)

  # Daily ltral information

  daily.data  <- read.csv(paste0(data.path, '/prepared/dailydata.csv'), header=T)

@@

  ltr.data <- read.csv(paste0(data.path, '/prepared/ltpropdata.csv'), header=T)

@@

  listing.data <- read.csv(paste0(data.path, '/prepared/ltlistdata.csv'), header=T)

@@

  str.keep.flds <- c('property.id', 'host.id', 'listing.title', 'property.type',
                     'listing.type', 'created.date', 'latitude', 'longitude',
                     'last.scraped.date', 'average.daily.rate', 'overall.rating',
                     'max.guests', 'cancellation.policy', 'cleaning.fee',
                     'published.nightly.rate', 'published.weekly.rate', 
                     'published.monthly.rate', 'minimum.stay',
                     'bedrooms', 'bathrooms', 'min.date', 'max.date', 'total.days',
                     'blocked.days', 'available.days', 'created.year')

@@

  ltr.keep.flds <- c('id.key', 'geographicalid', 'eventid', 'addressid', 'activityid',
                      'flatnumber', 'streetnumber', 'streetname', 'streettype', 
                      'suburb', 'postcode', 'property_latitude',
                      'property_longitude', 'street_centroid_latitude',
                      'street_centroid_longitude', 'eventdate', 'eventprice',
                      'propertytype', 'areasize', 'bedrooms', 'baths', 'parking',
                      'hasstudy', 'hascourtyard', 'hasbalcony', 'hasairconditioning',
                      'hasgarage', 'dom')

  ltr.data <- ltr.data[,ltr.keep.flds]

@@

  names(str.data) <- c('property.id', 'host.id', 'listing.title', 'property.type',
                       'listing.type', 'created.date', 'latitude', 'longitude',
                       'last.scraped.date', 'average.daily.rate', 'overall.rating',
                       'max.guests', 'cancel.policy', 'cleaning.fee',
                       'nightly.rate', 'weekly.rate', 'monthly.rate', 'min.stay',
                       'bedrooms', 'baths', 'min.date', 'max.date', 'total.days',
                       'blocked.days', 'available.days', 'created.year')

@@

>>>>

---
title: "Data Preparation, Integration and Cleaning"
output:
  pdf_document: default
  html_notebook: default
  html_document: default
---

@@

    data.path <- 'c:/dropbox/research/airBNB/data/'
    code.path <- 'c:/code/'
    fig.path <- 'c:/dropbox/research/airBNB/figures/'

@@

    data.path <- 'alternate data path'
    code.path <- 'alternate data path'
    fig.path <- 'alternate data path'

@@

Finally, we set the color to be used in the analysis.  We use AirBnb's official palette of 10 colors.

```{r set_colors, message=FALSE, warning=FALSE, comment=FALSE, echo=FALSE}

 # Set all colors

  abb.colors.names=c('rausch', 'hackberry', 'kazan', 'babu', 'lima', 'beach',
                     'ebisu', 'tirol', 'foggy', 'hof')
  abb.col <- c(rgb(250,88,99, maxColorValue=255), rgb(139,14,82, maxColorValue=255),
               rgb(0,117,140, maxColorValue=255), rgb(4,211,191, maxColorValue=255), 
               rgb(77,226,110, maxColorValue=255), rgb(252,179,14, maxColorValue=255),
               rgb(249,177,139, maxColorValue=255), rgb(191,166,100, maxColorValue=255),
               rgb(156,161,155, maxColorValue=255), rgb(86,94,97, maxColorValue=255))

  # Set specific colors

  str.col <- abb.col[1]
  ltr.col <- abb.col[3]

@@

  load(paste0(data.path, 'prepared/stpropdata.RData'))

  # Daily rental information

  load(paste0(data.path, 'prepared/stdailydata.RData'))

@@

  load(paste0(data.path, 'prepared/ltpropdata.RData'))

@@

  load(paste0(data.path, 'prepared/ltlistdata.RData'))

@@

  str.keep.flds <- c('property.id', 'host.id', 'listing.title', 'property.type',
                     'listing.type', 'created.date', 'latitude', 'longitude',
                     'average.daily.rate', 'overall.rating',
                     'max.guests', 'cancellation.policy', 'cleaning.fee',
                     'published.nightly.rate', 'published.weekly.rate', 
                     'published.monthly.rate', 'minimum.stay',
                     'bedrooms', 'bathrooms', 'min.date', 'max.date', 'total.days',
                     'block.rate', 'avail.rate', 'longest.block', 'nbr.block', 
                     'med.block', 'created.year')

@@

  ltr.keep.flds <- c('id.key', 'geographicalid', 'eventid', 'addressid', 'activityid',
                      'flatnumber', 'streetnumber', 'streetname', 'streettype', 
                      'suburb', 'postcode', 'property_latitude',
                      'property_longitude', 'street_centroid_latitude',
                      'street_centroid_longitude', 'event.date', 'eventprice',
                      'propertytype', 'areasize', 'bedrooms', 'baths', 'parking',
                      'hasstudy', 'hascourtyard', 'hasbalcony', 'hasairconditioning',
                      'hasgarage', 'dom')

  ltr.data <- ltr.data[ ,ltr.keep.flds]

@@

  names(str.data) <- c('property.id', 'host.id', 'listing.title', 'property.type',
                       'listing.type', 'created.date', 'latitude', 'longitude',
                       'average.daily.rate', 'overall.rating',
                       'max.guests', 'cancel.policy', 'cleaning.fee',
                       'nightly.rate', 'weekly.rate', 'monthly.rate', 'min.stay',
                       'bedrooms', 'baths', 'min.date', 'max.date', 'total.days',
                       'block.rate', 'avail.rate', 'longest.block', 'nbr.block', 
                       'med.block', 'created.year')

@@

### Trim by Time

Next we trim the observations by time.  Our hypothetical simulated decision is that of an investor who purchases a property on September 1, 2015 and must decide on traditional long term rental versus an AirBNB, short term rental approach to generating income.  To determine the likely long term rental rate for any given property, we will use the observed rental transactions from the market from September 1, 2014 to August 31, 2015.  Taking the AirBNB approach allows the owner to change rates over time and is subjected to daily changes in the market (supply and demand factors).  In this simulated example we will use AirBnB data from September 1, 2015 to August 31, 2016 to represent the actual AirBNB market conditions over the time period in question.  Another reason for using these two period of data is that the AirBNB data has missing daily values prior to September 1, 2015.  While we have imputed these missing daily observations, we hold the actual observed values to be of greater accuracy and have designed the initial simulated example to take advantage of the full observed set.

Before we begin the filter process, we initial the data cleaning counter -- a small function that keeps track of how many observations we remove.

```{r initialCC, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

  setCleanCount()

We then clip by time.  For the short-term data we will consider any Airbnb property that had at least 6 months worth of listing time on the portal during the 12-month study period

```{r clip_time, message=FALSE, warning=FALSE, comment=FALSE, cache=FALSE, echo=TRUE}

 # Limit short term data

  str.data <- str.data[(str.data$max.date - str.data$min.date) >= 178, ]
  str.data <- str.data[str.data$max.date >= '2016-03-01' &
                         str.data$min.date <= '2016-03-01', ]
  daily.data <- daily.data[daily.data$date >= '2015-09-01', ]

 # Limit long term data

  ltr.data <- ltr.data[ltr.data$event.date >= '2014-09-01' & 
                            ltr.data$event.date <= '2015-08-31', ]
  listing.data <- listing.data[listing.data$id.key %in% ltr.data$id.key, ]

 # Count cleaning

  countCleaning('time')

Next we add a monthly variable to the data for use in future modeling exercises.

```{r add_month, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

  # Short term data

  str.data$str.month <- 1
  daily.data$str.month <- as.numeric(as.factor(substr(daily.data$date, 1, 7)))

  # Long term data

  ltr.data$ltr.month <- as.numeric(as.factor(substr(ltr.data$event.date, 1, 7)))
  listing.data$ltr.month <- as.numeric(as.factor(substr(
    listing.data$event.date, 1, 7))) - 8

### Standardize Property Types

AirBNB units are classified into 19 different property type.  The most common are Apartments and Houses, the least common are Igloo, Tent, Treehouse and Yurt^[Property types are owner/lister defined, therfore it is likely that the Igloo, at the very least, is not a completely true representation of the property type]. For the sake of this analysis, we will collapse these twenty types into three types:

1. **House**:  Includes properties labeled 'House' or 'Townhouse'
2. **Apartment**: Includes properties labeled as 'Apartment' or 'Condominium'
3. **Other**: Includes properties labeled as 'Bed & Breakfast', 'Boat', 'Bungalow', 'Cabin', 'Camper/RV', 'Chalet', 'Dorm', 'Earth House', 'Hut', 'Igloo', 'Loft', 'Other', 'Tent', 'Treehouse', Villa' and 'Yurt.

Within the long-term rental data, property types fall into eight categories.  Like the AirBNB data we group these into three categories matching as best as possible to the AirBNB categories:

1. **House**:  Includes properties labeled 'Duplex', 'House', 'Terrace' or 'Townhouse'
2. **Apartment**: Includes properties labeled as 'Unit' or 'Studio'
3. **Other**: Includes properties labeled as 'Semi' and 'Villa'

Part of the difficulty in perfectly mapping property types from these two datasets is due to the fact that the long-term data uses Australian terms while the AirBNB data conforms to North American lexicon.  This is most apparent when talking about apartment dwellings.  In Australia these are referred to as Units, while in North American they are referred to as Apartments or Condominiums (depending on ownership structure).  Additionally, Terrace homes are very common in many of the older suburbs of Melbourne and would be considered Rowhouses or, likely, Townhomes in the North American context.

We then create a new variable in each dataset and give each one of the three categories discussed above.

```{r fix_type, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 # Short Term data

  str.data$type <- 'Other'
  str.data$type[str.data$property.type %in% c('Apartment', 'Condominium')] <- 'Apartment'
  str.data$type[str.data$property.type %in% c('House', 'Townhouse')] <- 'House'

 # Long Term data

  ltr.data$type <- 'Other'
  ltr.data$type[ltr.data$property.type %in% c('Duplex', 'House',
                                           'Terrace', 'Townhouse')] <- 'House'
  ltr.data$type[ltr.data$property.type %in% c('Studio', 'Unit')] <- 'Apartment'

As the chart below show, the **Other** category only makes up a very small percentage of the properties in both the short and long term market, especially in the long-term market.  Additionally, a good portion of the **Others** in the Airbnb data are Bed & Breakfast units, a use which doesn't fit with our research question.  As a result, we remove the 'Others' from both the Airbnb and the long term rental data.

```{r plot_type_comp, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE, fig.height=4}

  # Create Airbnb plot

  str.types <- ggplot(str.data, aes(x=type)) + 
    geom_bar(color=str.col, fill=str.col) +
    xlab('\nProperty Type\n') + 
    ggtitle('\nShort Term Rentals\n')  +
    ylab('# of Properties\n') +
    theme(plot.title = element_text(hjust = 0.5))

  # Create long term rental plot

  ltr.types <- ggplot(ltr.data, aes(x=type)) + 
    geom_bar(color=ltr.col, fill=ltr.col) +
    xlab('\nProperty Type\n') + 
    ggtitle('\nLong Term Rentals\n')  +
    ylab('# of Properties\n') +
    theme(plot.title = element_text(hjust = 0.5))

  # Output plots
  png(file=paste0(fig.path, 'dataclean_proptype.png'), width = 480, height = 280, 
      bg='transparent')
    ggMultiPlots(str.types, ltr.types, cols=2)
  dev.off()

```{r remove_others, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE, fig.height=4}

  str.data <- subset(str.data, type != 'Other')
  ltr.data <- subset(ltr.data, type != 'Other')

  # Count Cleaning
  countCleaning('property.type')

Next, for the short term rentals properties can be listed as one of three types relating to the extent of the property which is able to be booked:

1. **Entire Home/Apt**: The entire home or apartment is available
2. **Private Room**: One room within a house or apartment is available
3. **Shared Room**: A bed within a room shared by another occupant(s) is available

As long-term rentals do not offer **Private Room** or **Shared Room** options^[Our dataset does not include rooming houses, purpose built student accomodations and other types of rental which may offer private or shared room options] and our purpose here is a comparison of long and short term returnw, we remove all short term properties that do not lease the entire home/apt. Unfortunately, this filter does remove about 40% of the short term data, however, given the research question it is unavoidable.

```{r ltype_comp, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE, fig.height=4}

  # Fix data types and names
  str.data$listing.type <- as.character(str.data$listing.type)
  str.data$listing.type[str.data$listing.type == 'Entire home/apt'] <- 'Entire home'

  # Make Plot
  png(file=paste0(fig.path, 'dataclean_listtype.png'), width = 480, height = 280, 
      bg='transparent')

    ggplot(str.data, aes(x=listing.type)) + 
      geom_bar(color=str.col, fill=str.col) +
      xlab('\nListing Type\n') + 
      ggtitle('\nShort Term Rentals by Listing Type\n')  +
      ylab('# of Properties\n') +
    theme(plot.title = element_text(hjust = 0.5))

  dev.off()

```{r keep_entire_home, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

  str.data <- str.data[str.data$listing.type == 'Entire home', ]

  # Count cleaning 
  countCleaning('listing.type')

## Structural Characteristics

In this section we transform and filter based on the structural characteristics of the properties.

### Fix Boolean Variables

Before beginning, we fix a few of the boolean variables in the long term data that are stored in a non-compatible format.

```{r fix_lt_bools, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

  ltr.data$has.study <- ifelse(ltr.data$has.study == 'True', 1, 0)
  ltr.data$has.courtyard <- ifelse(ltr.data$has.courtyard == 'True', 1, 0)
  ltr.data$has.aircon <- ifelse(ltr.data$has.aircon == 'True', 1, 0)
  ltr.data$has.balcony <- ifelse(ltr.data$has.balcony == 'True', 1, 0)
  ltr.data$has.garage <- ifelse(ltr.data$has.garage == 'True', 1, 0)

### Bedrooms

We start by looking at bedrooms. We begin by removing all properties that are missing bedrooms information or have a likely data error in this fields (more than 14 bedrooms). We then plot the distribution of diffeltr bedrooms counts.  From this point onwards we bifurcate our analysis between houses and apartments as the markets (prices, rates, and properties) differ between the two property types.

```{r bed_fltr, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 # Remove missing and data errors

  # Long term data
  ltr.data <- ltr.data[!is.na(ltr.data$bedrooms), ]
  ltr.data <- ltr.data[ltr.data$bedrooms <= 10, ]

  # Short term data
  str.data <- str.data[!is.na(str.data$bedrooms), ]
  str.data <- str.data[str.data$bedrooms <= 10, ]

```{r bed_fltr2, message=FALSE, warning=FALSE, comment=FALSE, cache=FALSE, echo=TRUE, fig.height=5}

 # Build long term bar plot

   ltrbed <- ggplot(ltr.data, aes(x=bedrooms)) + 
    geom_bar(fill=ltr.col, color=ltr.col) +
    facet_wrap(~type,scales='free_y') +
    scale_x_continuous(labels=1:10, breaks=1:10)+
    xlab('\nBedrooms\n') +
    scale_y_continuous(labels=c(0,10000,20000), 
                       breaks=c(0,10000,20000))+
    ylab('') +
    ggtitle('Bedrooms in Long Term Rentals') +
    theme(plot.title = element_text(hjust = 0.5))

 # Build short term bar plot

  strbed <- ggplot(str.data, aes(x=bedrooms)) + 
    geom_bar(fill=str.col, color=str.col) +
    facet_wrap(~type,scales='free_y') +
    scale_x_continuous(labels=0:10, breaks=0:10) +
    xlab('\nBedrooms\n') +
    ylab('') +
    ggtitle('Bedrooms in Short Term Rentals') +
    theme(plot.title = element_text(hjust = 0.5))

  # Make Plot
  png(file=paste0(fig.path, 'dataclean_bedcount.png'), width = 480, height = 280, 
      bg='transparent')
    ggMultiPlots(strbed, ltrbed, cols=1)
  dev.off()

For both short and long term rentals we see that most apartments have less than 4 bedrooms and houses less than 5.  We filter the data accordingly. Additionally, note that short term rentals consider studio apartments to be 0 bed units while the long term data considers these as 1 bedroom. We convert the 0 bed short term units to 1 bed units.

```{r bed_fltr3, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 # Remove all with more than 4 beds

  str.data <- str.data[str.data$bedrooms < 5, ]
  ltr.data <- ltr.data[ltr.data$bedrooms < 5, ]

 # Remove apartments with more than 3

  str.id <- which(str.data$bedrooms > 3 & str.data$type =='Apartment')
  ltr.id <- which(ltr.data$bedrooms > 3 & ltr.data$type =='Apartment')

  if(length(str.id) > 0) str.data <- str.data[-str.id, ]
  if(length(ltr.id) > 0) ltr.data <- ltr.data[-ltr.id, ]

  # Convert 0 Bed to 1 Bed  
  str.data$bedrooms[str.data$bedrooms == 0] <- 1

  # Count Data Cleaning
  countCleaning('bedrooms')

### Bathrooms

We then turn to bathrooms, starting by looking at the distribution of baths between short/long and apt/house.

```{r bath_fltr, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 # Remove missing data

  ltr.data <- ltr.data[!is.na(ltr.data$baths), ]
  str.data <- str.data[!is.na(str.data$baths), ]

```{r bath_fltr2, message=FALSE, warning=FALSE, comment=FALSE, cache=FALSE, echo=TRUE, fig.height=5}

 # Build long term plot

  ltbath <- ggplot(ltr.data, aes(x=baths)) + 
    geom_bar(fill=ltr.col, color=ltr.col) +
    facet_wrap(~type) +
    scale_x_continuous(labels=1:8, breaks=1:8)+
    xlab('\nBathrooms\n') +
    ylab('') + 
    ggtitle('Bathrooms in Long Term Rentals') +
    theme(plot.title = element_text(hjust = 0.5))

 # Build short term plot

  stbath <- ggplot(str.data, aes(x=baths)) + 
    geom_bar(fill=str.col, color=str.col) +
    facet_wrap(~type, scales='free_y') +
    scale_x_continuous(labels=0:8, breaks=0:8)+
    xlab('\nBathrooms\n') +
    ylab('') + 
    ggtitle('Bathrooms in Short Term ltrals') +
    theme(plot.title = element_text(hjust = 0.5))

 # Make Plots  
  png(file=paste0(fig.path, 'dataclean_bathount.png'), width = 480, height = 280, 
      bg='transparent')

    ggMultiPlots(stbath, ltbath, cols=1)

  dev.off()

Here we see that nearly all properties have 3 or fewer bathrooms and we filter the data accordingly. A few short term properties state no baths, which is likely an error so we remove these as well. Also, here we notice that the short term data gives bathrooms in halves, while the long term data presents baths in whole numbers.  We round the baths up to whole integers.

```{r bath_fltr3, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 # Filter by bath count

  # Short term data
  str.data <- str.data[str.data$baths < 4, ]
  str.data <- str.data[str.data$baths > 0, ]

  # Long term data
  ltr.data <- ltr.data[ltr.data$baths < 4, ]

 # Round baths up

  str.data$baths <- round(str.data$baths + .01, 0)

  # Count cleaning
  countCleaning('baths')

### Bed/Bath Combination

In addition to looking at bedrooms and bathrooms as separate dimensions of properties, they can also be considered together.  In other words, it is very common in the industry (and when looking for a short term ltral) to specify searches by bed/bath combination for obvious reasons -- 1 bedroom with 3 baths is inefficient and 4 bedrooms with 1 bath is uncomfortable.  As a result we examine the combinations of the two.

```{r filtBB, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 # Create combination variable

  ltr.data$bedbath <- paste0(ltr.data$bedrooms, '..', ltr.data$baths)
  str.data$bedbath <- paste0(str.data$bedrooms, '..', str.data$baths)

```{r filtBB2, message=FALSE, warning=FALSE, comment=FALSE, cache=FALSE, echo=TRUE, fig.height=5}

 # Make long term plot

  ltbb <- ggplot(ltr.data, aes(x=bedbath)) + 
    geom_bar(fill=ltr.col, color=ltr.col) +
    xlab('\nBed/Bath Combo\n') +
    ylab('Number of Properties\n') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle('Long Term Rentals') + 
    facet_wrap(~type, scales='free_y')+
    theme(plot.title = element_text(hjust = 0.5))

 # Make short term plots

  stbb <- ggplot(str.data, aes(x=bedbath)) +
    geom_bar(fill=str.col, color=str.col) + 
    xlab('\nBed/Bath Combo\n') +
    ylab('Number of Properties\n') +
    ggtitle('Short Term Rentals') + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    facet_wrap(~type, scales='free_y')+
    theme(plot.title = element_text(hjust = 0.5))

 # Make Plots  
  png(file=paste0(fig.path, 'dataclean_bedbath.png'), width = 480, height = 280, 
      bg='transparent')

    ggMultiPlots(ltbb, stbb, cols=1)

  dev.off()

From this analysis we see that there are 6 combinations that make up most all properties:  1bed/1bath, 2/1, 2/2, 3/1, 3/2 and 4/2. We filter the data to these combinations.

```{r filtBB3, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 # Set level of acceptable combinations

  acc.bb <- c('1..1', '2..1', '2..2', '3..1', '3..2', '4..2')

 # Filter data

  ltr.data <- ltr.data[ltr.data$bedbath %in% acc.bb, ]
  str.data <- str.data[str.data$bedbath %in% acc.bb, ]

 # Count Cleaning

  countCleaning('bedbath')

Finally, we re-order the factor levels on the bed/bath combinations to make a 2-bed, 2-bath house or apartment the control case in future statistical models.  We additionally create a product variable that splits the bed/bath by apartment and house.

```{r message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

  # Re-order factors

  str.data$bedbath <- factor(str.data$bedbath, 
                              levels=c('2..2', '1..1', '2..1', '3..1', '3..2', '4..2'))
  ltr.data$bedbath <- factor(ltr.data$bedbath, 
                               levels=c('2..2', '1..1', '2..1', '3..1', '3..2', '4..2'))

  # Add product as a variable

  str.data$product <- paste0(substr(str.data$type, 1, 1), str.data$bedbath)
  ltr.data$product <- paste0(substr(ltr.data$type, 1, 1), ltr.data$bedbath)

## Rates and Rental values

We now turn to filtering by the prices -- the nightly rates and weekly rental values.  We start by removing observations missing price data.

```{r ltprice_fltr, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

  ltr.data <- ltr.data[!is.na(ltr.data$event.price), ]
  str.data <- str.data[!is.na(str.data$nightly.rate), ]

Next, we convert the Airbnb values to Australian dollars from US dollars using the exchange rate specified above (1.32).

```{r exch_rate_adj, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

  # Property level data

  str.data$average.daily.rate <- str.data$average.daily.rate * exch.rate
  str.data$nightly.rate <- str.data$nightly.rate * exch.rate
  str.data$weekly.rate <- str.data$weekly.rate * exch.rate
  str.data$monthly.rate <- str.data$monthly * exch.rate

  # Daily data

  daily.data$price <- daily.data$price * exch.rate

```{r ltprice_fltr2, message=FALSE, warning=FALSE, comment=FALSE, cache=FALSE, echo=TRUE}

  # ltr Plot

  ltr.plot <- ggplot(ltr.data, aes(x=event.price)) + 
    geom_density(fill=ltr.col, alpha=.8) +
    ylab('') +
    xlab('Weekly Rental') + 
    ggtitle('Long Term Rentals') + 
    scale_x_continuous(limits=c(0, 1500)) +
    theme(axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank()) +
    theme(plot.title = element_text(hjust = 0.5))

  # Nightly Rate plot
  rate.plot <- ggplot(str.data, aes(x=nightly.rate)) + 
    geom_density(fill=str.col, alpha=.8) + 
    ylab('') +
    xlab('Nightly Rate') +
    ggtitle('Short Term Rentals') + 
    scale_x_continuous(limits=c(0, 600)) +
    theme(axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank()) +
    theme(plot.title = element_text(hjust = 0.5))

```{r ltprice_fltr2s, message=FALSE, warning=FALSE, comment=FALSE, cache=FALSE, echo=TRUE}

  png(file=paste0(fig.path, 'dataclean_rates.png'), width = 480, height = 280, 
      bg='transparent')

    ggMultiPlots(ltr.plot, rate.plot, cols=2)

  dev.off()

From these density plots we see that most long-term Rentals are greater than $200 and less than $1000 per week while most nightly rates are greater than $50 and less than $500 per night. We filter the data accordingly.

```{r ltprice_fltr3, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 ## Filter the long term data

  ltr.data <- ltr.data[ltr.data$event.price >= 200 & 
                         ltr.data$event.price <= 1000, ]

 ## Filter the short term data

  str.data <- str.data[str.data$nightly.rate >= 50 &
                         str.data$nightly.rate <= 500, ]

  # Count Cleaning
  countCleaning('rates.rents')

### Locations and Submarkets

Location is clearly an important determinant in both the short and long term rental markets. At the broadest scale, the long-term Melbourne residential market is usually discussed in terms of Inner, Middle and Outer suburbs.  In general, prices are highest in the inner suburbs and lowest in the outer, with a number of exceptions in the high end neighborhoods in the east and southeastern areas of the middle suburbs.  Suburbs in Melbourne are much smaller than their North American counterparts.  The specific suburb of a property is the second, or finer scale, at which the market operates. The plot below shows the three large submarkets (Inner, Middle and Outer).

```{r mappoints, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 # Convert shapefile to ggplot-able object

  # Fix lat/long

  # Limit to extent of data
  sub.table <- table(ltr.data$suburb)
  sub.table <- sub.table[sub.table != 0]

  # Clip to only those with a ltral observation
  study.suburbs <- suburbs.shp[suburbs.shp@data$name_2006 %in% names(sub.table),]

  # Convert to an outer bound with holes
  ss.bound <- gUnaryUnion(study.suburbs)

  # Remove holes from the shapefile
  ssb <- slot(ss.bound, "polygons") 
  holes <- lapply(ssb, function(x) sapply(slot(x, "Polygons"), 
                                          slot, "hole"))
  res <- lapply(1:length(ssb), function(i) slot(ssb[[i]], "Polygons")[!holes[[i]]]) 
  IDs <- row.names(ss.bound) 
  ssb.fill <- SpatialPolygons(lapply(1:length(res), function(i) 
   Polygons(res[[i]], ID=IDs[i])), proj4string=CRS(proj4string(ss.bound)))

  # Identify which suburbs all in area covered by outer bound
  ss.in <- gIntersects(suburbs.shp, ssb.fill, byid=T)
  study.suburbs <- suburbs.shp[which(ss.in), ]

  # Convert into an outer boundary with no holes
  ss.bound.out <- gUnaryUnion(study.suburbs)
  ssb.out <- slot(ss.bound.out, "polygons") 
  holes <- lapply(ssb.out, function(x) sapply(slot(x, "Polygons"), 
                                          slot, "hole"))
  res.out <- lapply(1:length(ssb.out), 
                    function(i) slot(ssb.out[[i]], "Polygons")[!holes[[i]]]) 
  IDs <- row.names(ss.bound.out) 
  ssb.fill.out <- SpatialPolygons(lapply(1:length(res.out), 
                                         function(i) Polygons(res.out[[i]], 
                                                              ID=IDs[i])),
                                              proj4string=CRS(proj4string(ss.bound)))

  # Convert into a ggplot-able format
  subs <- fortify(study.suburbs, region='name_2006')

  # Add suburb category
  subs$des <- suburbs.shp@data$suburbdesi[
    match(subs$id, suburbs.shp@data$name_2006)]

```{r mappoints2, message=FALSE, warning=FALSE, comment=FALSE, cache=FALSE, echo=TRUE}

 ## Build the map object

  # Set colors
  g.cols <- abb.col[c(2, 10, 9)]

  # Make map
  loc.map <- ggplot() +
    geom_polygon(data=subs, aes(x=long, y=lat, group=group, fill=des), alpha=.8) + 
    scale_fill_manual(values=g.cols, labels=c('Inner      ',
                                              'Middle      ',
                                              'Outer')) +
    geom_path(data=subs, aes(x=long, y=lat, group=group), col='gray40') +
    scale_x_continuous(limits=c(min(subs$long), max(subs$long))) +
    scale_y_continuous(limits=c(min(subs$lat), max(subs$lat))) +
    xlab('') + ylab('')+
    theme(legend.position = 'bottom',
          legend.title = element_blank()) +
    ggtitle("Melbourne Suburb Classifications")+
    theme(plot.title = element_text(hjust = 0.5))

```{r plot_loc_map, message=FALSE, warning=FALSE, comment=FALSE, cache=FALSE, echo=TRUE, fig.height=6}
  png(file=paste0(fig.path, 'locationmap.png'), width = 480, height = 390, 
      bg='transparent')

    loc.map

  dev.off()

  loc.map

It is likely that the short-term rental market follows a similar market hierarchy, however, we can imagine that a few different spatial features can influence the short-term market.  While there are many possible additional features, we consider two to be the most likely to influence rates or occupancy: 1) Proximity to beach; and 2) Proximity to key tourist activities and events.  Using the three broad submarkets from the long-term market (inner, middle and outer) we have created a five sub-market system as a starting point for spatially analysing short and long term ltrals in Melboure:

1. Rural (Outer suburbs, not Beach)
2. Suburban (Middle suburbs, not Beach)
3. City (Inner Suburbs, not Beach, not Core)
4. City-Core (Select inner suburbs with tourist activities and near CBD)
5. Beach (Properties within 500m of Port Phillip Bay east of Yarra River)

We assign these designations by: 1) Adding suburb designations to the properties; 2) Assigning submarkets 1-4 based on suburb location; and 3) Indicating proximity to beach and labeling as 'Beach' submarket.  This process is repeated for both the short and long term data.  We also add the Statistical Area 1 code for each property for use in additional spatial modeling later on. Finally, we also remove any observations that fall outside of outer suburbs (the extent of the suburbs).

We select the 18 suburbs below to represent the 'core' of the city.  The majority of tourist destinations and major events such as the Australian Open and the Grand Prix are located in these suburbs.  All are well serviced by public transportation and possess abundant amenties for tourists.

```{r core_suburbs, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 core.suburbs <- c('Albert Park', 'Carlton', 'Collingwood', 'Cremorne', 'Docklands',
                   'East Melbourne', 'Fitzroy', 'Melbourne', 'Port Melbourne', 'Prahran',
                   'Richmond', 'South Melbourne', 'South Yarra', 'Southbank', 'St Kilda',
                   'St Kilda West', 'West Melbourne', 'Windsor')

We then apply a 500 meter buffer to the beach along Port Phillip Bay.

```{r beachbuffer, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 # Determine centroid of study area

  latMid <- median(str.data$latitude)
  longMid <- median(str.data$longitude)

 # Calculate Degress to Meters conversion

  m.lat = 111132.954 - 559.822 * cos(2 * latMid) + 1.175 * cos(4 * latMid)
  m.long = 111132.954 * cos (latMid)
  m.conv <- mean(m.lat, m.long)

 # Create the beach buffer

  beach.buffer <- gBuffer(beach.shp, byid=FALSE, width=500/m.conv)

Next we assign the suburb, SA1, city-core and beach designations to observations in the Airbnb dataset.

```{r add_st_space, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 ## Short term data

  # Convert to point shapefile
  str.prop.shp <- SpatialPointsDataFrame(cbind(str.data$longitude,
                                               str.data$latitude),
                                         str.data,
                                         proj4string=CRS("+init=epsg:4283"))

  # Add suburb name 
  spJoin <- over(str.prop.shp, suburbs.shp)
  str.data$suburb <- as.character(spJoin$name_2006)
  str.data$suburb[is.na(str.data$suburb)] <- 'missing'

  # Add suburb designation
  str.data$sub.class <- as.character(spJoin$suburbdesi)
  str.data$sub.class[is.na(str.data$sub.class)] <- 'missing'

  # Add SA1 designation
  spJoin <- over(str.prop.shp, sa1s.shp)
  str.data$sa1 <- spJoin$sa1_main11

  ## Add Beach Designation
  spJoin <- over(str.prop.shp, beach.buffer)
  str.data$beach <- ifelse(is.na(spJoin), 0 ,1)

  # Remove those missing a suburb classification (out of area)
  str.data <- str.data[str.data$sub.class != 'missing', ]
  str.data <- str.data[str.data$sa1 != 'missing', ]

```{r convert_st_subm, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 ## Add the submarket designations

  # Add individual classes
  str.data$sub.mrkt <- 'rural'
  str.data$sub.mrkt[str.data$sub.class == 'Middle'] <- 'suburban'
  str.data$sub.mrkt[str.data$sub.class == 'Inner'] <- 'city'
  str.data$sub.mrkt[str.data$suburb %in% core.suburbs] <- 'city-core'
  str.data$sub.mrkt[str.data$beach == 1] <- 'beach'

  # Order the factors properly
  str.data$sub.mrkt <- factor(str.data$sub.mrkt, 
                              levels=c('city-core', 'city', 'suburban', 
                                        'rural', 'beach'))

Before we do the same for the long term data, we must remedy the fact that about 5% of the long term rental observations have missing latitude and longitude values.  Half of these do, however, have a lat/long value for the centroid of the street that the property faces.  To retain as many observations as possible in these intial steps for those properties that lack a specific lat/long value we apply the street centroid value.  For those without either (3% of the total) we remove them from the dataset at this point.

```{r fix_ltt_geo, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 # Label those missing lat/long

  miss.LL <- which(is.na(ltr.data$latitude))

 # Create a field to indicate the approximate matches
